Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-08-31 paper-amia-summit-2018/amia_summit_2018.tex
fix reference format

\section*{Introduction}
Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics. The order of observations in a sequence is challenging to capture using features, which makes sequence classification a more challenging task than traditional classification. In general, sequence classification methods can be divided into three  categories: feature-based, distance metric-based and model-based method. Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}. Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well. Distance-based methods measure the similarity between sequences to determine the quality of classification. The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. The third type of sequence classification methods, first create a probabilistic model of a sequence, such as Markov chain and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).

However, Markov chain cannot handle long-term dependencies in sequences. Similarly, the hidden states of HMMs follow a first-order markov chain and can select only one hidden state at each time step. Therfore, markov chain and HMMs are only capable of passing short-term dependency among sequence steps. On the other hand, Recurrent Neural Networks have no such limitations. RNNs are capable to capture long-term dependencies with their unlimited memory. It was shown that RNNs are very effective in sequence modeling with temporal data.  Recently, two variant of RNNs are proposed: Long Short-Term Memory (LSTM)\cite{graves2013speech} and Gated Reccurent Unit (GRU)\cite{chung2014empirical}, which gave superior performance in a wide range of sequence modeling problems \cite{nion2013handwritten,lipton2015learning, choi2016doctor}.

@@

Our proposed baseline prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain (MC) and Hidden Markov Model (HMM), which are popular probabilistic models for discrete observation sequences with finite vocabulary. In the second step, we classified each test sequence based on the maximum likelihood of generating that sequence from each model. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}. However, the latest advances in deep learning technologis shows that deep learning, in particular RNNs, provide better results than conventional machine learning methods for the task of sequence classification. Therefore, we employed two special variant of RNN models in our experiemnt: long short-term memory (LSTM) and gated recurrent unit (GRU).

@@

\textbf {Recurrent Neural Networks (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit or simply GRU \cite{cho2014properties}. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from input sequence. Then, we applied a softmax layer at each time step for predicting the target. This target replication strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determine the relative importance of intermediate outputs. LSTM and GRU are trained on 70\% of the data and validated on 20\%. The remaining 10\% of the data is used as a test set for reporting the performance of the model. Our models contain several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

>>>>

\section*{Introduction}
Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics. The order of observations in a sequence is challenging to capture using features, which makes sequence classification a more challenging task than traditional classification. In general, sequence classification methods can be divided into three  categories: feature-based, distance metric-based and model-based method. Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}. Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well. Distance-based methods measure the similarity between sequences to determine the quality of classification. The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. The third type of sequence classification methods, first create a probabilistic model of a sequence, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).

However, Markov chain cannot handle long-term dependencies in sequences. Similarly, the hidden states of HMMs follow a first-order markov chain and can select only one hidden state at each time step. Therfore, markov chain and HMMs are only capable of passing short-term dependency among sequence steps. On the other hand, Recurrent Neural Networks (RNNs) have no such limitations. RNNs are capable to capture long-term dependencies with their unlimited memory. It was shown that RNNs are very effective in sequence modeling with temporal data.  Recently, two variant of RNNs are proposed: Long Short-Term Memory\cite{graves2013speech} (LSTM) and Gated Reccurent Unit\cite{chung2014empirical} (GRU), which gave superior performance in a wide range of sequence modeling problems \cite{nion2013handwritten, lipton2015learning, choi2016doctor}.

@@

Our proposed baseline prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain and Hidden Markov Model, which are popular probabilistic models for discrete observation sequences with finite vocabulary. In the second step, we classified each test sequence based on the maximum likelihood of generating that sequence from each model. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}. However, the latest advances in deep learning technologis shows that deep learning, in particular RNNs, provide better results than conventional machine learning methods for the task of sequence classification. Therefore, we employed two special variant of RNN models in our experiemnt: long short-term memory (LSTM) and gated recurrent unit (GRU).

@@

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit or simply GRU \cite{cho2014properties}. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from input sequence. Then, we applied a softmax layer at each time step for predicting the target. This target replication strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determine the relative importance of intermediate outputs. LSTM and GRU are trained on 70\% of the data and validated on 20\%. The remaining 10\% of the data is used as a test set for reporting the performance of the model. Our models contain several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.
